---
title: "05_analysis_1"
format: html
editor: visual
---

## Libraries

```{r}
#| message: false
#| warning: false
library("here")
library("tidyverse")

# Load custom functions if they exist
if(file.exists("99_proj_func.R")){
  source("99_proj_func.R")
}
```

## Load Data

```{r}
#| label: load-data

# NOTE: This block tries to load 03_augment first. 
# If 03 doesn't exist yet (because teammates haven't finished), it falls back to 02_clean.
# This ensures the document renders even if step 03 is missing.

path_03 <- here("data", "03_data_augment.tsv")
path_02 <- here("data", "02_data_clean.tsv")

if(file.exists(path_03)){
  message("Loading 03_data_augment.tsv...")
  experiment_data <- read_tsv(path_03, show_col_types = FALSE)
} else if(file.exists(path_02)){
  message("WARNING: 03_data_augment.tsv not found. Loading 02_data_clean.tsv instead.")
  experiment_data <- read_tsv(path_02, show_col_types = FALSE)
} else {
  stop("No data found! Please run 02_clean.qmd first.")
}
```

## Plot 1E (Group warm-up)

```{r}
# Everyone reproduces Fig 1E here - placeholder
```

## Plot 1F (Selina)

```{r}
# Placeholder for Selina
```

## Plot 1G (Marta)

```{r}
# Placeholder for Marta
```

## Plot 1H (Selina)

```{r}
# Placeholder for Selina
```

## Plot 1I (Tomasso)

```{r}
# Placeholder for Tomasso
```

## Plot 2A - Immunodominance landscape (Ivan)
Here we visualize the prevalence of T-cell responses for each peptide within specific HLA groups. We filter out HLA groups with fewer than 3 patients to ensure robustness.

```{r}
#| label: plot-2a-ivan
#| fig-width: 10
#| fig-height: 6

# --- 1. Prepare Data for Plot 2A ---
# Create a local copy to avoid modifying the global 'experiment_data' for others
df_2a <- experiment_data

# Standardize column names if necessary (based on 02_clean structure)
if("sampleID" %in% names(df_2a)) df_2a <- rename(df_2a, patient_id = sampleID)
if("Protein" %in% names(df_2a))  df_2a <- rename(df_2a, ORF = Protein)
if("Peptide" %in% names(df_2a))  df_2a <- rename(df_2a, peptide = Peptide)

# Ensure binary response exists (if 03_augment is not ready, we calculate it from p-value)
# p < 0.05 is considered a positive response
if(!"response_binary" %in% names(df_2a)) {
  df_2a <- df_2a %>% mutate(response_binary = if_else(p < 0.05, 1, 0))
}

# --- 2. Calculate Prevalence ---

# A. Count patients per HLA (distinct patients only)
hla_counts <- df_2a %>%
  distinct(patient_id, HLA) %>%
  count(HLA, name = "n_patients")

# B. Filter HLAs with < 3 patients
valid_hlas <- hla_counts %>%
  filter(n_patients >= 3) %>%
  pull(HLA)

# C. Summarize data: calculate prevalence per Peptide within HLA
plot_data <- df_2a %>%
  filter(HLA %in% valid_hlas) %>%
  group_by(HLA, peptide, ORF) %>%
  summarise(
    total_tested = n_distinct(patient_id),
    responders = sum(response_binary, na.rm = TRUE),
    prevalence = responders / total_tested,
    .groups = "drop"
  ) %>%
  filter(prevalence > 0) # Keep only positive responses for cleaner plot

# --- 3. Generate Plot ---

ggplot(plot_data, aes(x = peptide, y = ORF)) +
  geom_point(aes(size = prevalence, color = HLA), alpha = 0.7) +
  scale_size_continuous(range = c(2, 8), name = "Prevalence") +
  theme_bw() +
  labs(
    title = "SARS-CoV-2 T-cell Immunodominance Landscape",
    subtitle = "Response prevalence per HLA allele (n >= 3 patients)",
    x = "Peptide Epitope",
    y = "Viral ORF"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    legend.position = "right"
  )
